Finding the Optimal Coefficient Forms

### 2024-02-13 src/explore/20240213-optcoefs.qmd
It looks like we can estimate the variance. This holds if we just use X1 and X2 or if we use f(X1) and f(X2).

---
title: "Using GLS and Estimating the Covariance Matrix"
author:
  - name: Caleb Leedy
date: today
date-format: "D MMMM YYYY"
format:
  pdf:
    include-in-header: 
      - "latex_header.tex"
---

```{r}
#| label: global variables
#| echo: false
B <- 1000

obs_seg <- 250
true_mu <- 5
cov_e1e2 <- 0.5

@@

In this document, we

1. Propose a new technique for estimating a covariance matrix, and 
2. Show via simulation studies that it works.

# Problem and Proposal

## Problem

Previously, have a model $\hat g = Z g + e$ where $\hat g$ is defined by

\begin{align*}
g_1^{(11)} &= n^{-1} \sum_{i = 1}^n \frac{\delta_{11}}{\pi_{11}} g_1(x_i) \\
g_2^{(11)} &= n^{-1} \sum_{i = 1}^n \frac{\delta_{11}}{\pi_{11}} g_2(y_{1i}) \\
g_3^{(11)} &= n^{-1} \sum_{i = 1}^n \frac{\delta_{11}}{\pi_{11}} g_3(y_{2i}) \\
g_1^{(10)} &= n^{-1} \sum_{i = 1}^n \frac{\delta_{10}}{\pi_{10}} g_1(x_i) \\
g_2^{(10)} &= n^{-1} \sum_{i = 1}^n \frac{\delta_{10}}{\pi_{10}} g_2(y_{1i}) \\
g_1^{(01)} &= n^{-1} \sum_{i = 1}^n \frac{\delta_{01}}{\pi_{01}} g_1(x_i) \\
g_3^{(01)} &= n^{-1} \sum_{i = 1}^n \frac{\delta_{01}}{\pi_{01}} g_2(y_{2i}) \\
g_1^{(00)} &= n^{-1} \sum_{i = 1}^n \frac{\delta_{00}}{\pi_{00}} g_1(x_i) \\
\end{align*}

and

$$\hat g = 
\begin{bmatrix}
g_1^{(11)} \\
g_2^{(11)} \\
g_3^{(11)} \\
g_1^{(10)} \\
g_2^{(10)} \\
g_1^{(01)} \\
g_3^{(01)} \\
g_1^{(00)} \\
\end{bmatrix},
Z = 
\begin{bmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1 \\
1 & 0 & 0 \\
0 & 1 & 0 \\
1 & 0 & 0 \\
0 & 0 & 1 \\
1 & 0 & 0 \\
\end{bmatrix},
E[e] = 0,
\text{ and } 
\Var(e) = n^{-1}
\begin{bmatrix}
V_{11} & 0 & 0 & 0 \\
0 & V_{10} & 0 & 0 \\
0 & 0 & V_{01} & 0 \\
0 & 0 & 0 & V_{00} \\
\end{bmatrix}.
$$

We also have

$$
V_{11} = 
\begin{bmatrix}
\frac{1}{\pi_{11}} E[g_1^2] - E[g_1]^2 & \frac{1}{\pi_{11}} E[g_1g_2] -
E[g_1]E[g_2] & \frac{1}{\pi_{11}} E[g_1g_3] - E[g_1]E[g_3] \\
\frac{1}{\pi_{11}} E[g_1g_2] - E[g_1]E[g_2] & \frac{1}{\pi_{11}} E[g_2^2] -
E[g_2]^2 & \frac{1}{\pi_{11}} E[g_2g_3] - E[g_2]E[g_3] \\
\frac{1}{\pi_{11}} E[g_1g_3] - E[g_1]E[g_3] & \frac{1}{\pi_{11}} E[g_2g_3] -
E[g_2]E[g_3] & \frac{1}{\pi_{11}} E[g_3^2] - E[g_3]^2 \\
\end{bmatrix},
$$
$$
V_{10} = 
\begin{bmatrix}
\frac{1}{\pi_{10}}E[g_1^2] - E[g_1]^2 & \frac{1}{\pi_{10}}E[g_1g_2] - E[g_1]E[g_2]\\
\frac{1}{\pi_{10}}E[g_1g_2] - E[g_1]E[g_2] & \frac{1}{\pi_{10}}E[g_2^2] - E[g_2]^2 \\
\end{bmatrix},
$$
$$
V_{01} = 
\begin{bmatrix}
\frac{1}{\pi_{01}}E[g_1^2] - E[g_1]^2 & \frac{1}{\pi_{01}}E[g_1g_3] - E[g_1]E[g_3]\\
\frac{1}{\pi_{01}}E[g_1g_3] - E[g_1]E[g_3] & \frac{1}{\pi_{01}}E[g_3^2] - E[g_3]^2 \\
\end{bmatrix}, \text{ and }
V_{00} = 
\begin{bmatrix}
\frac{1}{\pi_{00}}E[g_1^2] - E[g_1]^2
\end{bmatrix}.
$$

However, this challenge with actually using this model is that we have to use
the matrix $V$, which we assume to be *known*. In this write up, we will not
make this assumption.

## Proposal

Instead of using the known covariance matrix $V$, we estimate it instead.
Suppose that we observe variables $Z = (X_{m_1}, Y_{m_2})$ with an objective of
estimating $\theta = E[g(Y_{m_2})$ for some known function $g$ such that the
variables $Z$ are subject to missingness. We assume that there are $R$ unique
combinations of observed variables including a fully observed case. We can index
the combinations of observed variables by $r$ and assume that the fully observed
case occurs at $r = 1$. Let $G_r(Z)$ be the variables that are observed at a
particular value of $r$. We can choose a sequence of functions $f_1, \dots, f_K$
that we want to estimate. Each $f_k$ is assumed to be a function of a subset of
$Z$ and it makes sense to assume (since these are chosen by the analyst) that
there is at least one $f_k$ for each combination of observed variables $G_r(Z)$.
Let $A_k$ be the sets of observed variable combinations that can be evaluate by
the function $f_k$. If $f_k$ can be evaluated by the observed variables
$G_r(Z)$, this will consist of all of the combinations of variables
$r'$ such that $G_r(Z) \subseteq G_r'(Z)$. Because we assume that $G_1(Z) = Z$,
$A_k$ is always non-empty.

To estimate the covariance matrix $V$, we estimate the covariance between 
$f_{k_1}$ and $f_{k_2}$ directly by computing the estimated covariance on 
$A_{k_1} \cap A_{k_2}$.

<!---# This did not work because the orthogonalization created variables that we
needed to impute
### Proposal 2: Orthogonal g-functions

For this idea we choose functions that after estimation are approximately
independent from each
other with a standard variance. If this is true, the covariance matrix $V$ is
approximately the identity matrix. Not only do we not have to estimate $V$, but
we do not even need to invert it! I now give a more detailed explanation of this
method.

Suppose that we observe variables $Z = (X_{m_1}, Y_{m_2})$ with an objective of
estimating $\theta = E[g(Y_{m_2})$ for some function $g$ such that the
variables are subject to missingness. We assume that there are $R$ unique
combinations of observed variables including a fully observed case. We can index
the combinations of observed variables by $r$ and assume that the fully observed
case occurs at $r = 1$. Let $G_r(Z)$ be the variables that are observed at a
particular value of $r$. We can choose a sequence of functions $f_1, \dots, f_K$
that we want to estimate. Each $f_k$ is assumed to be a function of a subset of
$Z$ and it makes sense to assume (since these are chosen by the analyst) that
there is at least one $f_k$ for each combination of observed variables $G_r(Z)$.
Let $A_k$ be the sets of observed variable combinations that can be evaluate by
the function $f_k$. If $f_k$ can be evaluated by the observed variables
$G_r(Z)$, this will consist of all of the combinations of variables
$r'$ such that $G_r(Z) \subseteq G_r'(Z)$. Because we assume that $G_1(Z) = Z$,
$A_k$ is always non-empty.

Once we have the original $f_1, \dots, f_K$ we orthogonalize them using a
Gram-Schmidt process. Let $g_1 = f_1 / \hat \Var(f_1)$. Then for $k > 1$
we can continue using the sequence,

$$\tilde g_k = f_k - \sum_{i = 1}^k \frac{\hat \Cov(f_k, g_i)}{\hat \Var(f_k)}
g_i \text{ followed by } g_k = \tilde g_k / \hat \Var(\tilde g_k).$$

However, the parameters for $\tilde g_k$ and $k > 1$ are all regression
coefficients and the multiplier for $g_k$ is simply computing a variance.
Then to
achieve efficiency, we propose estimating the regression coefficients by running
the coresponding regression on all of the data points in $A_k$ as well as
computing the variance of $\tilde g_k$ within $A_k$.
--->

@@

<!---# Compare with: Known V in non-monotone case --->
## Non-Monotone Case

First, we consider the following non-monotone simulation setup.

\begin{align*}
  \begin{bmatrix} x_1 \\ e_1 \\ e_2 \end{bmatrix} 
  &\stackrel{ind}{\sim} N\left(\begin{bmatrix} 0 \\ 0 \\ 0 \end{bmatrix}, 
  \begin{bmatrix} 1 & 0 & 0 \\ 0 & 1 & \rho \\ 0 & \rho & 1 \end{bmatrix}\right) \\
  x_2 &= x + e_1 \\
  y &= \mu + x + e_2 \\
\end{align*}

This yields outcome variables $X_1$ and $X_2$ that are correlated both with $Y$
and additionally with each other. To generate the missingness pattern, we select
`r obs_seg` observations into the four segments independently using simple
random sampling. The goal of this estimation problem is to estimate $\theta = E[Y]$.

@@

```{r}
#| label: proposed estimation
#| echo: false

@@

    return(as.numeric(t(c(0, 0, 1) %*% ghat)))
  } else {
    stop("Only class == 'basic' has been implemented.")
  }
}

@@

```{r}
#| label: MCMC
#| echo: false
#| tbl-cap: |
#|   Results from simulations study with independent equally sized segments
#|   $A_{11}$, $A_{10}$, $A_{01}$, and $A_{00}$ all of size $n = 250$. In this
#|   simulation we have the true mean of $Y_2$ equal to $\mu = 5$ and the covariance
#|   between $e_1$ and $e_2$ is $\rho = 0.5$. The goal is to estimate 
#|   $E[Y_2] = \mu$. For the GLS estimation, we use the
#|   estimated covariance matrix $\hat V$ with 
#|   f-functions $f_1 = X_1$, $f_2 = X_2$ and $f_3 = Y$.

@@

    gls <- gls_est(df, rho = cov_e1e2)
    gls_ev <- est_var_gls(df)

    return(tibble(Oracle = oracle,
                  CC = cc,
                  IPW = ipw,
                  GLS = gls,
                  GLSEstVar = gls_ev
    ))

@@

<!---# Compare with: Best case in monotone data 
## Monotone Case--->

# Optimal choice of g-functions

TODO for the next set of simulations.

# Conclusion

Did it work? Yes, estimating the covariance matrix did not produce a loss in
efficiency.

>>>>

---
title: "Finding the Optimal Coefficient Forms"
author:
  - name: Caleb Leedy
date: today
date-format: "D MMMM YYYY"
format:
  pdf:
    include-in-header: 
      - "latex_header.tex"
---

```{r}
#| label: global variables
#| echo: false

B <- 2000
obs_seg <- 250
true_mu <- 5
cov_e1e2 <- 0.5

@@

In this document, we do the following:

1. Vary the functional forms of the coefficients, 
2. Check to ensure that our estimated variance is still accurate.

# Idea

In `20240202-estcovmat.qmd`, we tested an estimated covariance matrix. The results
suggest that estimating the covariance matrix suffers little penalty. In this
report, we want to see if choosing different functional forms of $f_1$ and $f_2$ 
(that depend on the data) cause the covariance estimation to be biased.

Instead of using $f_1(X_1) = X_1$ and $f_2(X_2) = X_2$ as we did in 
`20240202-estcovmat.qmd` we want to use $f_1(X_1) = E[Y \mid X_1]$ and 
$f_2(X_2) = E[Y \mid X_2]$, *and* we need to also test them with estimated
versions of these functions.

@@

@@

```{r}
#| label: estimated variance basic
#| echo: false

@@

    } else {
    stop("Only class == 'basic' has been implemented.")
  }
}

@@

```{r}
#| label: estimated variance opt
#| echo: false

estvar_glsopt <- function(df, class = "opt") {

  if (class == "opt") {

    z_mat <- matrix(rep(1, 8), ncol = 1)

    # Get Regression Estimates
    mx1 <- lm(Y ~ X1, data = filter(df, delta_11 == 1 | delta_01 == 1))
    mx2 <- lm(Y ~ X2, data = filter(df, delta_11 == 1))
    df <- df |>
      mutate(mx1 = predict(mx1, newdata = df)) |>
      mutate(mx2 = predict(mx2, newdata = df))

    # Put Estimates into g_hat
    g_hat <- 
      c(mutate(df, tmp = delta_11 / pi_11 * mx1) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_11 / pi_11 * mx2) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_11 / pi_11 * Y) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_10 / pi_10 * mx1) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_10 / pi_10 * mx2) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_01 / pi_01 * mx1) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_01 / pi_01 * Y) |> pull(tmp) |> mean(),
        mutate(df, tmp = delta_00 / pi_00 * mx1) |> pull(tmp) |> mean())

    # Estimate Covariance Matrix
    df_1p <- filter(df, delta_11 == 1 | delta_10 == 1)
    df_p1 <- filter(df, delta_11 == 1 | delta_01 == 1)
    df_11 <- filter(df, delta_11 == 1)
    v_x1 <- var(df$mx1)
    v_x2 <- var(df_1p$mx2)
    v_y <- var(df_p1$Y)
    cov_x1x2 <- cov(df_1p$mx1, df_1p$mx2)
    cov_x1y <- cov(df_p1$mx1, df_p1$Y)
    cov_x2y <- cov(df_11$mx2, df_11$Y)

    return(as.numeric(ghat))
  } else {
    stop("Only class == 'opt' has been implemented.")
  }
}

```{r}
#| label: MCMC
#| echo: false
#| tbl-cap: |
#|   Results from simulations study with independent equally sized segments
#|   $A_{11}$, $A_{10}$, $A_{01}$, and $A_{00}$ all of size $n = 250$. In this
#|   simulation we have the true mean of $Y_2$ equal to $\mu = 5$ and the covariance
#|   between $e_1$ and $e_2$ is $\rho = 0.5$. The goal is to estimate 
#|   $E[Y_2] = \mu$. For the GLS estimation, we use the
#|   estimated covariance matrix $\hat V$ with 
#|   f-functions $f_1 = X_1$, $f_2 = X_2$ and $f_3 = Y$. For the GLS optimal
#|   estimation we are trying $f_1 = E[Y \mid X_1]$, $f_2 = E[Y \mid X_2]$ and 
#|   $f_3 = Y$. Each of these expectations are estimated with using linear
#|   regression. So $E[Y \mid X_i]$ is implemented as a regression estimator
#|

@@

    gls <- gls_est(df, rho = cov_e1e2)
    gls_ev <- est_var_gls(df)
    gls_opt <- estvar_glsopt(df)

    return(tibble(Oracle = oracle,
                  CC = cc,
                  IPW = ipw,
                  GLS = gls,
                  GLSEstVar = gls_ev,
                  GLSOpt = gls_opt
    ))

@@

```
# Conclusion

Is the covariance estimation still working? (By working we mean causing the
estimator to be unbiased.)

* It looks like estimating the variance is not causing additional bias in the
  estimator. This is great. 
* We may, however, need to check more complicated simulation setups because the
  simpicity of the current setup might make it look better than it actually is.
* One suprising thing is that the "optimal" estimator is not better than the
  other estimators. This could be because of the simplicity of the setup, but we
  may need to check it out.
